---
title: "SSL"
author: "Alia Shahzad"
date: "5/14/2020"
output: github_document


---

```{r setup, echo = FALSE}
#The necessary packages

library(tidyverse)
library(here)
install.packages("RSocrata") #Make sure you have this one!
library(forcats)

#Sends all the visualizations to a folder named images (and creates that folder if it 
#doesn't exist). I might use this folder to fetch images for my final write-up.

knitr::opts_chunk$set(
  fig.path = "images/"
)


```
## Introduction

This report uses  publically available data from the Strategic Subject List (SSL), street name the "Heat List," which was retired in 2019. During its time, the Chicago Police Department (CPD) used an algorithm to assess the risk of a given individual becoming party to violence. The CPD intended to anticipate and intervene with subjects on the SSL before potential violence occured. But the SSL was subject to much public scrutiny, one reason being the lack of transparancy around the algorithm and the data being fed into it, and it's use was quietly retired in 2019.

Under pressure from the ACLU, the state of ILlinois' Attorney General's Office, as well as a group of journalists who submitted a FOIA for SSL data, in 2016, the CPD released a set of de-identified data of list subjects that included, among other categories, scores, race, age arrested, and current age; these are the variables I analyze in this report. The CPD [maintained](https://chicagounbound.uchicago.edu/cgi/viewcontent.cgi?article=1636&context=uclf) that it did not use race as a factor when calculating risk scores. While it's impossible to assess the truth of this claim as they declined to release the SSL algorithm itself, this analysis also aims to determine to what extent risk scores were assosciated by race. 

SSL Scores are placed on a scale ranging from 0 (extremely low risk of violence) to 500 (extremely high risk violence). A score above 250 [earns](https://medium.com/equal-future/how-strategic-is-chicagos-strategic-subjects-list-upturn-investigates-9e5b4b235a7c) "heightened police scrutiny," and score above 450 is high-risk for becoming party to violence. The vast majority of scores are between 0 - 450, and within this group, more scores are high risk (250-350) than low risk (0-250). Very few scores are very high risk (above 450).

```{r read & tidy data, echo = TRUE, results = 'hide', cache = TRUE}

#Builds local file path & avoids using the dreaded setwd() by using Socrata package
#to read directly from a URL. Find more info on Socrata.  [here](https://github.com/Chicago/RSocrata/blob/master/README.md).

ssl <- RSocrata::read.socrata("https://data.cityofchicago.org/Public-Safety/Strategic-Subject-List/4aki-r3np/data") %>%
  write_csv(path = here("data", "ssl.csv")) %>% #writing data into right folder
  select(ssl_score, 
         predictor_rat_age_at_latest_arrest,
         predictor_rat_gang_affiliation, 
         race_code_cd, 
         age_curr) %>% #selecting vars of interest
  rename(age_arr = 'predictor_rat_age_at_latest_arrest',
         gang_aff = 'predictor_rat_gang_affiliation', 
         race = 'race_code_cd') #shortening var names


```
  
      
```{r ssl split factors, echo = TRUE, results = 'hide'}

#this chunk splits the ssl scores into new classes for the next graph 
#it's in a different chunk so I can hide its very ugly product when it print

mylabels <- c("Low Risk", "High Risk", "Very High Risk") 
cut_ssl <- cut(ssl$ssl_score, breaks=c(0, 250, 450, 500), labels=mylabels)  
  
```

## SSL Scores Overview by Risk of Violence


```{r ssl scores overview graph, results = FALSE}

#plotting data visualization
ssl_overview <- ssl %>%
  #mapping ssl_scores & cut ssl_scores separately to show score groups in visualization
  ggplot(mapping = aes(ssl_score, fill = cut_ssl)) + 
  geom_area(stat = "bin") +
  labs(title = "SSL Score Distribution", 
       x = "SSL Score",
       y = "Count",
       fill = "Risk of Becoming Party to Violence",
       caption = "(Source: Chicago Data Portal)") +
  theme(legend.position = "top")
  theme_minimal() 

ssl_overview
```

SSL Scores are placed on a scale ranging from 0 (extremely low risk of violence) to 500 (extremely high risk violence). A score above 250 [earns](https://medium.com/equal-future/how-strategic-is-chicagos-strategic-subjects-list-upturn-investigates-9e5b4b235a7c) "heightened police scrutiny," and score above 450 is high-risk for becoming party to violence. From this graph, we observe that the vast majority of scores are between 0 - 450, and within this group, more scores are high risk (250-350) than low risk (0-250). Very few scores are at very high risk (above 450).

## SSL By Age

```{r scores by age}

age_arr_mut <- ssl %>%
  #using forcats:fct_relevel to re-order levels to desired order along x axis
  mutate(age_arr = fct_relevel(age_arr, 
                               c("less than 20",
                                 "20-30",
                                 "30-40",
                                 "40-50",
                                 "50-60",
                                 "60-70",
                                 "70-80"))) %>% 
  #using forcats:fct_collapse to combine older age groups in over fifty
  mutate(age_arr = fct_collapse(age_arr,
                                'over fifty' = c("50-60", 
                                                 "60-70", 
                                                 "70-80"))) 
age_arr_bar <- age_arr_mut %>%
  #dropping na's before graphing
  drop_na() %>%
  ggplot(mapping = aes(age_arr)) +
  #adding a fun color
  geom_bar(fill = "#FF6666") +
  labs(title = "SSL Subjects by Age Last Arrested", 
             x = "Age Group",  
             y = "Count",
             caption = "(Source: Chicago Data Portal)") +
  theme_minimal()


age_arr_bar



ssl_age_curr <- ssl %>% 
  #using forcats:fct_relevel to re-order levels to desired order along x axis
  mutate(age_curr = fct_relevel(age_curr, 
                               c("less than 20",
                                 "20-30",
                                 "30-40",
                                 "40-50",
                                 "50-60",
                                 "60-70",
                                 "70-80"))) %>% 
  #using forcats:fct_collapse to combine older age groups in over fifty
  mutate(age_curr = fct_collapse(age_curr,
                                'over fifty' = c("50-60", 
                                                 "60-70", 
                                                 "70-80")))
ssl_age_group <- ssl_age_curr %>%
  #get rid of na's before graphing
  drop_na() %>%
  ggplot(mapping = aes(x = ssl_score, fill = age_curr, na.rm = FALSE)) +
  geom_density(stat = "bin") +
   labs(title = "SSL Subjects by Current Age Group", 
             x = "SSL Score",  
             y = "Density",
             fill = "Age Group", 
             caption = "(Source: Chicago Data Portal)") +
  theme_minimal()
  

ssl_age_faceted <- ssl_age_curr %>%
  #get rid of na's before graphing
  drop_na() %>%
  ggplot(mapping = aes(x = ssl_score, na.rm = FALSE)) +
  geom_density() +
  facet_wrap(~age_curr) +
  labs(title = "SSL Subjects by Current Age Group, Faceted", 
             x = "SSL Score",  
             y = "Density",
             caption = "(Source: Chicago Data Portal)") +
  theme_minimal()
  

ssl_age_group
ssl_age_faceted
```
From the first graph, which counts SSL subjects by age first arrested, we observe that most SSL subjects were under the age of 40 when they were most recently arrested. The most represented age group of most recent arrest is from 20-30, which is unsurprising considering that this is also the age group most represented in committing crimes in the city. For further analysis, it might be interesting to compare these figures to age groups of SSL subjects when they were first arrested, but unfortunately the CPD did not release this data. I suspect we'd see a significantly larger bar for the under 20 age group in this case, since many of the subjects are arrested repeatedly throughout their lives. 

From the second and third grapsh, which depict current age of Subjects, we note that age group with the highest scores is under 20! As subjects get older, their risk for becoming party to violence decreases. These observations lines up with age demographics of violent crime offenders; past offenders are significantly less likely to commit crimes as they age, particularly coming out of long periods of incarceration.

## Accuracy and Age

```{r oldies visualization}

ssl_oldies <- ssl %>%
  group_by(ssl_score) %>%
  #count total number of SSL scores per score
  mutate(total = n()) %>%
  #count (oldies) for old high risk SSL subjects
  filter(age_curr == "50-60" | 
           age_curr == "60-70" | 
           age_curr == "70-80" & 
           ssl_score > 250) %>% 
  mutate(oldies = n()) %>% 
  #create percentage
  mutate(percent = oldies/total) %>% 
  #graph
  ggplot(mapping = aes(x = ssl_score, y = percent)) +
  geom_line() +
  # print the y-axis labels using percentages rather than proportions
  scale_y_continuous(labels = scales::percent) +
  # label our graph
  labs(
    title = "Percent of Old Subjects (Age 50+)  per SSL Score",
    x = "SSL Score",
    y = "Percent of Age 50+ Subjects",
    caption = "Source: Chicago Data Portal"
  )

ssl_oldies

```

A common critique of the SSL was that its input data was not updated frequently enough, leading to the assignment of high scores to elder individuals, who are significantly less violence-prone that people under 50. In other words, keeping people on the list who had last been arrested when they were in their 20s but are now 40 or 50 might lead to inaccurate data. In this graph, I look at what percent of old people (current aged 50+) constitute each SSL score bracket. The graph demonstrates that people over the age of 50 are assesed at a fairly low-risk score (below 250), which indicates that this worry is likely not founded. We would be more worried if the elderly composed any percentage of the scores above 250.

## Gang Affiliation

```{r age & gang affiliation}

ssl_age_gang <- ssl_age_curr %>%
  #get rid of na's before graphing
  drop_na() %>%
  #reordering by bar height and filling by gang affiliation
  ggplot(mapping = aes(x = age_curr, 
                       fill = (gang_aff == TRUE), 
                       na.rm = FALSE)) +
  geom_histogram(stat = "count") +
   labs(title = "SSL Subjects Gang Affiliation by Age Group", 
             x = "Age Group",  
             y = "Count of Subjects",
             fill = "Gang Affiliation", 
             caption = "(Source: Chicago Data Portal)") +
  #manually customizing colors and legend text
  scale_fill_manual(values = c("#E69F00", "#56B4E9"), 
                       name = "Gang Affiliation",
                       breaks = c("TRUE", "FALSE"),
                       labels = c("Affiliated", "Not Affiliated")) 
  
#next creating a boxplot of gang affiliations and ssl scores  
#first creating a list for factor labels
  gang_labels <- list(
  '0' = "No Gang Affiliation",
  '1' = "Gang Affiliation"
)
#creating a function to push labels (I know this is extra work but trying
#to make up for the lack of tidying I had to do).

gang_labeller <- function(variable,value){
  return(gang_labels[value])
}

ssl_age_gang_box <- ssl_age_curr %>%
  #get rid of na's before graphing
  drop_na() %>%
  #reordering by bar height and filling by gang affiliation
  ggplot(mapping = aes(x = ssl_score, 
                      na.rm = FALSE)) +
  geom_boxplot() +
  #faceting around gang affilliation 
  facet_wrap(~gang_aff,
             labeller = gang_labeller) +
  labs(title = "SSL Subjects Gang Affiliation by Age Group", 
             x = "SSL Score",  
             caption = "(Source: Chicago Data Portal)")

ssl_age_gang
ssl_age_gang_box

```
Here, we examine the relationship between SSL scores, age and gang affiliation. In the faceted boxplot, we observe that the median SSL score is a bit higher for people with gang affiliations, but maybe not as much as we might anticipate. In the bar graph, we observe that subjects currently aged 20-30 are most likely to be members of gangs. This is unsurprising given the trend of younger criminals (20 and below) to opt out of gangs and instead participate in smaller cliques or rogue acts of violence.

## SSL By Race

```{r by race}

ssl_race_data <- ssl %>% 
  #using forcats:fct_relevel to re-order levels to desired order along x axis
  mutate(race = fct_collapse(race,
                             'Other' = c("I", "U", "WBH"),
                             'Black' = c("BLK"), 
                             'White' = c("WHI"),
                             'Hispanic' = c("WWH")))


ssl_race_bar <- ssl_race_data %>% 
  #reordering by bar height and filling by gang affiliation
  ggplot(mapping = aes(x = race, fill = fct_infreq(race))) +
  geom_bar() +
   labs(title = "SSL Subject Counts by Race", 
             x = "Race Group",  
             y = "Counts",
             caption = "(Source: Chicago Data Portal)") +
  theme_minimal() +
  theme(legend.title = element_blank())

ssl_race_box <- ssl_race_data %>% 
  #reordering by bar height and filling by gang affiliation
  ggplot(mapping = aes(x = ssl_score)) +
  geom_boxplot() +
  facet_wrap(~race)
   labs(title = "SSL Subject Counts Faceted by Race", 
             x = "Race Group",  
             y = "Count of Subjects",
             caption = "(Source: Chicago Data Portal)") +
  theme_minimal() +
  theme(legend.title = element_blank())

ssl_race_bar
ssl_race_box

```

These graphs map SSL scores by race. Looking to the boxplots, we observe no significant differences in median SSL score by race. However, in the bar charts, we do see that significantly more black men are subject on the list than any other race, at a proportion signicantly higher than their perctentage of the city's population, [30%](https://www.census.gov/quickfacts/chicagocityillinois). 


